---
title: "A basic semicompeting risks model in Stan"
author: "Leah Comment"
date: "10/25/2017"
output:
  pdf_document: default
header-includes:
  - \usepackage{tikz}
  - \usetikzlibrary{positioning}
  - \usetikzlibrary{arrows}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Model formulation

## Weibull models

The hazard of a Weibull model with shape $\alpha$ and scale $\kappa$ is
\[ h(t|x_i) = \alpha \kappa t^{\alpha-1} \exp\left\lbrace x_i'\beta \right\rbrace\]
This is the parameterization used in Lee et al.

Alternatively, one can use the Stan hazard parameterization, where
\[ h(t) = \frac{\alpha}{\sigma} \left( \frac{t}{\sigma} \right)^{\alpha - 1}\]

The relationship between the two is that 
\[ \sigma^{-\alpha} = \exp\left\lbrace x_i'\beta \right\rbrace\]

# Likelihood contributions by observed data pattern

## Observed

The Stan code to fit these models is shown below.
```{r, code = readLines("semicompeting_weibull.stan"), eval = FALSE}
```

```{r, cache = TRUE}
# Fit Weibull semicompeting model
fit1 <- stan(file = "semicompeting_weibull.stan", data = dat_ID, iter = 500, chains = 1)
```

