---
title: "A basic semicompeting risks model in Stan"
author: "Leah Comment"
date: "10/25/2017 (updated 10/28/2017)"
output:
  pdf_document: default
header-includes:
  - \usepackage{tikz}
  - \usepackage{xcolor}
  - \usepackage{pgfplots}
  - \usetikzlibrary{positioning}
  - \usetikzlibrary{arrows}
---

\definecolor{hcolor}{RGB}{179,226,205}
\definecolor{icolor}{RGB}{255,242,174}
\definecolor{dcolor}{RGB}{253,205,172}
\definecolor{rcolor}{RGB}{51,160,44}
\definecolor{tcolor}{RGB}{31,120,180}
\definecolor{dscolor}{RGB}{27,158,119}
\definecolor{cscolor}{RGB}{217,95,2}

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Illness-death models

## Introduction
The compartmental model representation of the issue of semicompeting risks conceptualizes an observational unit as belonging to exactly one state at any given time point. These models apply to contexts outside of illness and death, but it is illustrative to think of a person who may be "healthy," "ill," or "dead." Developing illness is a non-terminal event and dying is the terminal event. When illness occurs, it _must_ occur prior to death. However, some subjects may die without ever experiencing the illness of interest.

\begin{figure}[ht!]
\centering
\begin{tikzpicture}[ ->,shorten >=2pt,>=stealth, node distance=2cm, pil/.style={->,thick,shorten =2pt,}, box/.style = {fill=#1,
                 draw, solid, thin, minimum size=9mm}]
	\node[box = hcolor] (h) {\textbf{Healthy}};
	\node[box = icolor, above right=of h] (i) {\textbf{Ill}};
	\node[box = dcolor, below right=of i] (d) {\textbf{Dead}};
	\path[->] (h) edge node[below]{$\lambda_2$} (d);
    \path[->] (h) edge node[above]{$\lambda_1 \ \ $} (i);
    \path[->] (i) edge node[above]{$\ \ \lambda_3$} (d);
\end{tikzpicture}
\caption{A compartmental model representation of an illness-death model in semicompeting risks}
\label{fig:compart}
\end{figure}

Making inferences in the illness-death context involves modeling the compartment transition hazards $\lambda_1$, $\lambda_2$, and v$\lambda_3$. These transitions hazards can change as a function of time, yielding $\lambda_1(t)$, $\lambda_2(t)$, and $\lambda_3(t)$.

## Notation

Suppose we have a non-terminal event $R$ and a terminal event $T$. The terminal event can be any absorbing state, but "terminal event" and "death" are used interchangeably here. Both event times are subject to a censoring time which may vary by person. Let $T_i$ denote the terminal event time for person $i$. When the non-terminal event occurs, $R_i \in \mathbb{R}_+$; when the non-terminal event is truncated by death, $R_i \equiv \bar{\mathbb{R}}$, a non-real number. The potential censoring time is $C_i$. The observed non-terminal event time is $Y^R_i = \min(R_i, T_i, C_i)$, with $\min(x, \bar{\mathbb{R}}) \equiv x$ for any $x \in \mathbb{R}$. The non-terminal event indicator $\delta^R_i$ is $\mathbf{1}(Y^R_i = R_i)$. Similarly, $Y^T_i = \min(T_i, C_i)$, and $\delta^R_i$ is $\mathbf{1}(Y^T_i = T_i)$.

## Potential observed data patterns

Defined by $V_i = (\delta^R_i, \delta^T_i)$, four distinct data patterns are possible:

| Type | Description                                            | $(Y^R_i, Y^T_i)$ | $V_i = (\delta^R_i, \delta^T_i)$ |
|:----:|--------------------------------------------------------|:------------------------:|:--------------------------:|
| 1    | Observe neither non-terminal nor terminal event        | $(c_i, c_i)$     | $(0,0)$                    |
| 2    | Observe non-terminal event; terminal event is censored | $(r_i, c_i) = (y^R_i, c_i)$   | $(1,0)$                    |
| 3    | Observe terminal event without non-terminal event      | $(t_i, c_i) = (y^T_i, y^T_i)$ | $(0,1)$                    |
| 4    | Observe non-terminal and terminal events               | $(r_i, t_i) = (y^R_i, y^T_i)$ | $(1,1)$                    |


## Defining the joint distribution of non-terminal and terminal event times

The joint density for $T$ and $R$ is shown in Figure \ref{fig:jointdens}. The portion where both event types are observed is only defined on the "upper wedge" where $T>R$; the joint density is $f_u(r,t)$. When the non-terminal event does not occur ($R = \bar{\mathrm{R}}$), the density of $T$ is give $f_{\bar{\mathrm{R}}}(t)$.

\begin{figure}[h]
\label{fig:jointdens}
\begin{center}
\begin{tikzpicture}
\pgfplotsset{ticks=none}
\begin{axis}[title = {$f_u(r,t)$},
scale only axis = true, enlargelimits = true,
width = 0.3\linewidth, height = 0.3\linewidth,
axis lines = left, xlabel = {Non-terminal event time $R$}, 
x label style = {at = {(axis description cs:0.5,-0.05)}, anchor = north},
ylabel = {Terminal event time $T$}]
\addplot[domain=0:1, samples=3, color=black] {x};
\addplot[fill = lightgray, draw = none, fill opacity = 0.25] coordinates {(0,0) (1,1) (0,1)};
\node at (axis cs:0.25, 0.65) {$T>R$};
\end{axis}
\end{tikzpicture}
\begin{tikzpicture}
\pgfplotsset{ticks=none}
\begin{axis}[title = {$f_{\bar{\mathbb{R}}}(t)$},
scale only axis = true, enlargelimits = true, 
width = 0.1\linewidth, height = 0.3\linewidth,
axis lines = left, xlabel = {$R = \bar{\mathbb{R}}$}, ylabel = {},
x axis line style = {draw = none}, 
x label style = {at = {(axis description cs:0,-0.05)}, anchor = north}]
\addplot[domain=0:2.2, samples=100, color=black] (1.5 * x^0.5 * exp(-x^1.5), x);
\end{axis}
\end{tikzpicture}
\end{center}
\end{figure}

For now, we assume a semi-Markov model for $\lambda_3$ where the hazard of the terminal event only depends on covariates and time elapsed since the non-terminal event. Ignoring measure theoretic issues and using $P(X=x)$ to refer to the pdf $f(x)$ for any continuous random variable,

\begin{align*} 
f_u(r,t) & = P(R = r, T = t) \text{ for } t > r \\
& = P(R = r, T > r) P (T = t | R = r, T > r) \\
& = 
\underbrace{\frac{P(R = r, T > r)}{\textcolor{dscolor}{P(R > r, T > r)}}}_{\lambda_1(r)}
\underbrace{\textcolor{dscolor}{P(R > r, T > r)}}_{S_1(r) S_2(r)}
\underbrace{\frac{P(T = t | R = r, T > r)}{\textcolor{cscolor}{P(T > t | R = r)}}}_{\lambda_3(t-r)}
\underbrace{\textcolor{cscolor}{P(T > t | R = r)}}_{S_3(t-r)} \\
& = \lambda_1(r) \lambda_3(t-r) \exp\left\lbrace - \Lambda_1(r) - \Lambda_3(t-r) \right\rbrace
\end{align*}

In a Markov model for death following the non-terminal event, $\lambda_3(t-r)$, $\Lambda_3(t-r)$, and $S_3(t-r)$ above would be replaced by $\lambda_3(t)$, $\Lambda_3(t)$, and $S_3(t)$.

The density for terminal event times among those who never experience the non-terminal event is
\begin{align*}
f_{\bar{\mathbb{R}}} & = P(T = t, R = \bar{\mathbb{R}}) \\
& = 
\underbrace{\frac{P(R > t, T = t)}{\textcolor{dscolor}{P(R > r, T > r)}}}_{\lambda_2(t)}
\underbrace{\textcolor{dscolor}{P(R > r, T > r)}}_{S_1(t)S_2(t)} \\
& = \lambda_2(t) \exp \left\lbrace -\Lambda_1(t) + \Lambda_2(t) \right\rbrace
\end{align*}

## Likelihood contributions by data pattern

For the semi-Markov model, the general form of the likelihood contributions for each observed data pattern in the table above is

\[ (\textcolor{rcolor}{\text{non-terminal contribution}}) \times 
(\textcolor{tcolor}{\text{terminal contribution}}) \]

Type 1 (observe neither event type):
\[ L_{1i} = P(R > c_i, T > c_i) 
= \textcolor{rcolor}{S_1(c_i)} \textcolor{tcolor}{S_2(c_i)} 
= \textcolor{rcolor}{S_1(y^R_i)} \textcolor{tcolor}{S_2(y^T_i)} 
\]

Type 2 (observe only non-terminal event):
\[ L_{2i} = P(R = y^R_i, T > c_i) = \int_{c_i}^\infty f_u(y^R_i, u) du
= \textcolor{rcolor}{f_1(y^R_i)} \textcolor{tcolor}{S_2(y^R_i) S_3(c_i - y^R_i)}
= \textcolor{rcolor}{f_1(y^R_i)} \textcolor{tcolor}{S_2(y^R_i) S_3(y^T_i - y^R_i)}
\]

Type 3 (observe only terminal event):
\[ L_{3i} = P(R > y^T_i, T = y^T_i) = \int_{y^T_i}^\infty f_u(u, y^R_i) du 
= \textcolor{rcolor}{S_1(y^T_i)} \textcolor{tcolor}{f_2(y^T_i)}
\]

Type 4 (observe both event types):
\[ L_{4i} = P(R = y^R_i, T = y^T_i) = f_u(y^R_i, y^T_i)
= \textcolor{rcolor}{f_1(y^R_i)} 
\textcolor{tcolor}{S_2(y^R_i) f_3(y^T_i - y^R_i)}
\]

## Full likelihood

The combined likelihood of the observed data for $n$ observations is
\begin{align*}
\mathcal{L}_n = &
%1
\left[ \prod_{i : (\delta^R_i, \delta^T_i)=(0,0)}  
\textcolor{rcolor}{S_1(y^R_i)} \textcolor{tcolor}{S_2(y^T_i)} 
\right]
%2
\left[ \prod_{i : (\delta^R_i, \delta^T_i)=(1,0)} 
\textcolor{rcolor}{f_1(y^R_i)} \textcolor{tcolor}{S_2(y^R_i) S_3(y^T_i - y^R_i)}
\right] \times \\
& 
%3
\left[ \prod_{i : (\delta^R_i, \delta^T_i)=(0,1)} 
\textcolor{rcolor}{S_1(y^T_i)} \textcolor{tcolor}{f_2(y^T_i)}
\right]
%4
\left[ \prod_{i : (\delta^R_i, \delta^T_i)=(1,1)} 
\textcolor{rcolor}{f_1(y^R_i)} 
\textcolor{tcolor}{S_2(y^R_i) f_3(y^T_i - y^R_i)} 
\right]
\end{align*}

TODO(LCOMM): fill in notation gaps above


# Model formulation

## Weibull models

The hazard of a Weibull regression model with shape $\alpha$ is
\[ h(t|x_i) = \alpha t^{\alpha-1} \exp\left\lbrace \beta_0 + x_i'\beta \right\rbrace\]
With $\kappa = \exp(\beta_0)$, this is the parameterization used in Lee et al.

Alternatively, one can use the Stan hazard parameterization, where
\[ h(t) = \frac{\alpha}{\sigma} \left( \frac{t}{\sigma} \right)^{\alpha - 1} \]

To incorporate regression parameters into the model, we note that
\[ \sigma^{-\alpha} = \exp\left\lbrace x_i'\beta \right\rbrace \]
and thus
\[ \sigma = \exp\left\lbrace -\frac{x_i'\beta}{\alpha} \right\rbrace \]

# Fitting semicompeting risk Weibull hazard models in Stan

Suppose we use Weibull hazards for the illness-death model below, which includes only a single covariate (`x_c` in the `dat_ID` data set):

Non-terminal event hazard at time $r$:
\[ h_{1}(r | X_i) 
= \alpha_1 \kappa_1 r^{\alpha_1 - 1} \mathrm{exp} \left\lbrace \beta_1 X_i \right\rbrace
= \alpha_1 r^{\alpha_1 - 1} \mathrm{exp} \left\lbrace \beta_0 + \beta_1 X_i \right\rbrace 
\text{ for } r > 0 \]

Terminal event hazard at time $t$, without having experienced the non-terminal event:
\[ h_{2}(t | X_i) 
= \alpha_2 \kappa_2 t^{\alpha_2 - 1} \mathrm{exp} \left\lbrace \beta_2 X_i \right\rbrace
= \alpha_2 t^{\alpha_2 - 1} \mathrm{exp} \left\lbrace \beta_{02} + \beta_2 X_i \right\rbrace 
\text{ for } t > 0 \]

Terminal event hazard at time $t$, after having experienced the non-terminal event at $r$ in a Semi-Markov model:
\[ h_{3}(t | X_i) 
= \alpha_3 \kappa_3 (t-r)^{\alpha_3 - 1} \mathrm{exp} \left\lbrace \beta_3 X_i \right\rbrace 
= \alpha_3 (t-r)^{\alpha_3 - 1} \mathrm{exp} \left\lbrace \beta_{03} + \beta_3 X_i \right\rbrace 
\text{ for } t > r \]

Let $\mathbf{x_i'\boldsymbol{\beta}}_g = \beta_{0g} + \beta_{g} X_i$ for $g \in (1,2,3)$. (This is the linear predictor that appears in the Stan code.)


```{r}
# Packages
library("knitr")
suppressPackageStartupMessages(library("rstan"))
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores() - 1)
```

```{r}
# Read in semicompeting risk Weibull simulated data
dat_ID <- readRDS("dat_ID.Rdata")

# Package for Stan
x_m <- cbind(1, dat_ID$x_c)
stan_dat <-  list(N = nrow(dat_ID),
                  P_1 = ncol(x_m),
                  X_1 = x_m,
                  P_2 = ncol(x_m),
                  X_2 = x_m,
                  P_3 = ncol(x_m),
                  X_3 = x_m,
                  Yr = dat_ID$R,
                  dYr = dat_ID$delta_R,
                  Yt = dat_ID$T,
                  dYt = dat_ID$delta_T)

```

The Stan code to fit these models is shown below.
```{r, code = readLines("semicompeting_weibull.stan"), eval = FALSE}
```

```{r, cache = TRUE, fig.height = 3, fig.width = 5, fig.align = "center"}
# Fit Weibull semicompeting model
library("rstan")
# Discards first half of chains -> will yield 2000 posterior samples total
fit1 <- stan(file = "semicompeting_weibull.stan", data = stan_dat, 
             iter = 1000, chains = 4)
s_elapsed <- sum(get_elapsed_time(fit1))
plot(fit1)
print(fit1)
```

TODO(LCOMM): Look into reparameterization to reduce numerical issues

```{r, include = FALSE}
# Compare to Lee et al
library("SemiCompRisks")

# Objects
Y = data.frame(dat_ID$R, dat_ID$delta_R, 
               dat_ID$T, dat_ID$delta_T)
lin.pred = list(formula(~ x_c),
                formula(~ x_c),
                formula(~ x_c))


# Start lightly adapted code from SemiCompRisks documentation -------------------

### Hyperparameters
## Subject-specific frailty variance component
##  - prior parameters for 1/theta
theta.ab <- c(0.7, 0.7)

## Weibull baseline hazard function: alphas, kappas
WB.ab1 <- c(0.5, 0.01) # prior parameters for alpha1
WB.ab2 <- c(0.5, 0.01) # prior parameters for alpha2
WB.ab3 <- c(0.5, 0.01) # prior parameters for alpha3
##
WB.cd1 <- c(0.5, 0.05) # prior parameters for kappa1
WB.cd2 <- c(0.5, 0.05) # prior parameters for kappa2
WB.cd3 <- c(0.5, 0.05) # prior parameters for kappa3

hyperParams <- list(theta = theta.ab,
                    WB = list(WB.ab1 = WB.ab1, WB.ab2 = WB.ab2, WB.ab3 = WB.ab3,
                              WB.cd1 = WB.cd1, WB.cd2 = WB.cd2, WB.cd3 = WB.cd3))

###################
## MCMC SETTINGS ##
###################

## Setting for the overall run
# Get 2000 total post-burnin reps after combining 4 chains
# 500 samples per chain -> 5000 samples pre-thinning -> 10000 samples/chain
numReps    <- 10000
thin       <- 10
burninPerc <- 0.5

## Settings for storage
nGam_save <- 0
storeV    <- rep(TRUE, 3)

## Tuning parameters for specific updates
mhProp_theta_var  <- 0.05
mhProp_Vg_var     <- c(0.05, 0.05, 0.05)

## Specific to the Weibull specification of the baseline hazard functions
mhProp_alphag_var <- c(0.01, 0.01, 0.01)
mcmc.WB  <- list(run = list(numReps = numReps, thin = thin, burninPerc = burninPerc),
                 storage = list(nGam_save = nGam_save, storeV=storeV),
                 tuning = list(mhProp_theta_var = mhProp_theta_var,
                               mhProp_Vg_var = mhProp_Vg_var, 
                               mhProp_alphag_var = mhProp_alphag_var))

## Chain seed starting values
myModel <- c("semi-Markov", "Weibull")
startValues      <- vector("list", 4)
startValues[[1]] <- initiate.startValues_HReg(Y, lin.pred, dat_ID, model = myModel)
startValues[[2]] <- initiate.startValues_HReg(Y, lin.pred, dat_ID, model = myModel,
                                              theta = 0.23)
startValues[[3]] <- initiate.startValues_HReg(Y, lin.pred, dat_ID, model = myModel,
                                              theta = 0.72)
startValues[[4]] <- initiate.startValues_HReg(Y, lin.pred, dat_ID, model = myModel,
                                              theta = 0.55)

# End lightly adapted code from SemiCompRisks documentation ---------------------

```

We use the `SemiCompRisks` package to fit the illness-death model. The model fit by `BayesID_HReg()` is not completely comparable to the Stan code above because it includes an individual-level $\gamma$ frailty that induces additional correlation between $R_i$ and $T_i$. (However, the Stan code could be changed to include a frailty parameter.)

The models fit by `SemiCompRisks` are as follows:

Non-terminal event hazard at time $r$:
\[ h_{1}(r | \gamma_i, X_i) 
= \gamma_i \alpha_1 \kappa_1 r^{\alpha_1 - 1} \mathrm{exp} \left\lbrace \beta_1 X_i \right\rbrace 
= \gamma_i \alpha_1 r^{\alpha_1 - 1} \mathrm{exp} \left\lbrace \beta_0 + \beta_1 X_i \right\rbrace 
\text{ for } r > 0 \]

Terminal event hazard at time $t$, without having experienced the non-terminal event:
\[ h_{2}(t | \gamma_i, X_i) 
= \gamma_i \alpha_2 \kappa_2 t^{\alpha_2 - 1} \mathrm{exp} \left\lbrace \beta_2 X_i \right\rbrace 
= \gamma_i \alpha_2 t^{\alpha_2 - 1} \mathrm{exp} \left\lbrace \beta_{02} + \beta_2 X_i \right\rbrace 
\text{ for } t > 0 \]

Terminal event hazard at time $t$, after having experienced the non-terminal event at $r$ in a Semi-Markov model:
\[ h_{3}(t | \gamma_i, X_i) 
= \gamma_i \alpha_3 \kappa_3 (t-r)^{\alpha_3 - 1} \mathrm{exp} \left\lbrace \beta_3 X_i \right\rbrace 
= \gamma_i \alpha_3 (t-r)^{\alpha_3 - 1} \mathrm{exp} \left\lbrace \beta_{03} + \beta_3 X_i \right\rbrace 
\text{ for } t > r \]

The $\gamma_i$ are assumed to be i.i.d. from a $\Gamma(\theta^{-1}, \theta^{-1})$ distribution, which has a mean of 1. The models assumed by the Stan code can be seen as a special case where $\gamma_i = 1 \ \forall \ i$.

```{r}
# Fit SemiCompRisks models
k_start <- Sys.time()
fit2 <- BayesID_HReg(Y, lin.pred, data = dat_ID,
                     hyperParams = hyperParams, startValues = startValues,
                     mcmc = mcmc.WB, path = "SemiCompRisks/Output/")
k_elapsed <- Sys.time() - k_start
```

The Stan model took `r round(s_elapsed, 2)` seconds and the `SemiCompRisks` model took `r round(k_elapsed, 2)` seconds to run. Although the models are not exactly comparable, we can compare our estimates from these two fits, especially for $(\beta_1, \beta_2, \beta_3)$.

TODO(LCOMM): Check on these times (expes)
TODO(LCOMM): Get effective sample size/PSRF numbers to compare 2 approaches.

```{r}
# Comparing estimates from SemiCompRisks and Stan models

# SemiCompRisks posterior means? medians? can't tell 
# TODO(LCOMM): Find this out
k_theta_est <- summary(fit2)[["theta"]][,1]
k_kappa_ests <- exp(summary(fit2)[["h0"]][1, c(1,4,7)])
k_beta0_ests <- summary(fit2)[["h0"]][1, c(1,4,7)]
k_alpha_ests <- exp(summary(fit2)[["h0"]][2, c(1,4,7)])
k_beta1_ests <- log(summary(fit2)[["coef"]][c(1,4,7)])
semicomp_param_ests <- c(k_theta_est, k_beta0_ests, k_beta1_ests, k_alpha_ests)

# Stan posterior means
s_pmeans <- summary(fit1)[["summary"]][,"mean"]
s_beta0_ests <- s_pmeans[c("beta1[1]", "beta2[1]", "beta3[1]")]
s_beta1_ests <- s_pmeans[c("beta1[2]", "beta2[2]", "beta3[2]")]
s_alpha_ests <- s_pmeans[c("alpha1", "alpha2", "alpha3")]
names(s_beta0_ests) <- c("$\\beta_{01}$", "$\\beta_{02}$", "$\\beta_{03}$")
names(s_beta1_ests) <- c("$\\beta_{1}$", "$\\beta_{2}$", "$\\beta_{3}$")
names(s_alpha_ests) <- c("$\\alpha_{1}$", "$\\alpha_{2}$", "$\\alpha_{3}$")
stan_param_ests <- c(s_beta0_ests, s_beta1_ests, s_alpha_ests)

# Combine into a table
library("knitr")
semi_res_tab <- data.frame(Parameter = c("$\\theta$", names(stan_param_ests)),
                           stan = c(NA, stan_param_ests),
                           semicomp = semicomp_param_ests)
options(knitr.kable.NA = "--")
kable(semi_res_tab, escape = FALSE, row.names = FALSE, align = "crr",
      col.names = c("Parameter", "Estimate from Stan", "Estimate from SemiCompRisks"),
      caption = "Illness-death parameter estimates", digits = 4)
```


TODO(LCOMM): Make posterior density plot comparison (maybe?)

TODO(LCOMM): Some kind of convergence comparison -> PSRF?

TODO(LCOMM): Implement Stan model with frailty

